Uncertainty in the evolution of social learning

### 2022-05-12 ms.tex
updated figures and analysis text

\newcommand{\meansl}{\langle s \rangle}

@@

We developed an agent-based model of a society of $N$ individuals who each must decide
which of $B$ behaviors to perform at each time step. Each behavior is a ``bandit'',
a common modelling and experimental approach for representing behaviors with
probabilistic payoffs~\cite{SuttonBartoBook, McElreath2005,Rendell2010,Schulz2019}. %Bandits are like  slot machines in a casino~\ref{fig:behaviorSelection}.
Each behavior $b$ yields a payoff of 1 with probability $\pi_b$ and a payoff of zero
otherwise; $\pi_b$ is therefore also the expected payoff of behavior $b$. Agents must
decide which behavior to perform at each time step. To do this, agents 
use an explore-exploit strategy to sometimes try the most profitable behavior
they know about, and other times try alternatives that may pay off more reliably.

We operationalized uncertainty in four different ways: 
(1) payoff ambiguity, $A_\pi$, which measures the difference
between the optimal expected payoff behavior $\pisub{high}$ and the expected payoff
of the other behaviors, $\pisub{low}$; 
(2) environmental variability, $u$, the probability the optimal behavior changes from one generation to another; 
(3) the number of possible behaviors, $B$; and 
(4) the lifespan, or time steps per generation, $M$.

We ran the simulation for $T=1000$ time steps. At
the end of each generation agents reproduce and then die off. 
Those selected to reproduce
pass on their boolean social learning trait, $s$, without mutation,
which specifes whether agents learn socially. We developed a series of
computational analyses where we systematically vary the uncertainty
variables and observe the average prevalence of social learning trait $s$
across agents and trials, which is our primary outcome measure, 
denoted $\langle s \rangle$.

@@

In each time step, $N$ agents---autonomous problem solvers---select 
which behavior to perform based on their running tally of mean payoffs for 
each available behavior. %information (Figure~\ref{fig:behaviorSelection}). 
Agent $i$ tracks the mean payoff of each behavior $b$, denoted $\bar\pi_{ib}$, and
a \emph{count} of how many times it has performed each behavior, denoted $c_{ib}$.
$\bar\pi_{ib}$
is initialized to 0 for all $b$ at model initialization for
all agents, and after each generation for individual learner agents. Social
learning agents' $\bar\pi_{ib}$ is initialized as that of its teacher selected
through performance-biased oblique learning. Agent $i$'s accumulated payoffs
from performing several behaviors over their lifetime of $M$ time steps is
denoted $\pi_{i}$.

\vspace{2em}

\begin{table}[h]
    \caption{Dynamic agent-level variables. Each has an implicit time dependence.}
    \label{tab:modelParameters}
    \centering \hspace{-3em}
    \begin{tabular}{cp{4.0in}l} \toprule

@@

        $\pi_i$ & Net payoffs accumulated by $i$ within generation
                                & 0.0 \\
        \bottomrule
    \end{tabular}
\end{table}

\vspace{2em}

\subsection{Modeling uncertainty}

In our model uncertainty is a tunable consequence of four environmental features:
%that structure the behavior of agents.  
% Many factors contribute to the 
% uncertainty involved in both social and asocial learning. 
%We model uncertainty in four distinct ways: 
\begin{enumerate}
    \item We vary the frequency at which the optimal behavior changes, which we call \emph{environmental variability}, $u$. At the start of each generation, with probability $u$, a new behavior is assigned a payoff of $\pisub{high}$ while all other behaviors are assigned a payoff of $\pisub{low}$. Otherwise, the same behavior remains optimal across generations. 
    \item We vary the latent expected payoffs yielded by the bandits. For simplicity,
      we assume that in any given environmental state, there is one optimal behavior that
      yields an expected payoff of $\pisub{high}$, while all other behaviors yield a 
      payoff of $\pisub{low} < \pisub{high}$. As $\pilow$ increases towards 
      $\pihigh$, uncertainty in the form of \emph{payoff ambiguity} also increases,
      making it more difficult, and less rewarding, for agents to determine which
      behavior is the optimal one.
      % The difference between these quantities is the \emph{payoff ambiguity}, 
    % \begin{equation}
    %   A_\pi = 1 - (\pisub{high} - \pisub{low}),
    %   \label{eq:payoffAmbiguity}
    % \end{equation}
    % which is maximized when the expected payoffs are close to equal and minimized
    % when they differ greatly. For simplicity we will set $\pisub{high} = 0.9$
    % for all analyses and vary $\pisub{low}$ in order to vary $A_\pi$.
    \item We vary the total \emph{number of available behaviors}, $B$, which is a
      source of uncertainty since agents are less likely to know which behavior yields high payoffs.  
    \item Finally, we vary number of behavioral events per generation, $M$. This can be viewed as the \emph{effective lifespan} of an agent.  Decreasing this lifespan effectively increases the importance of each event for acquiring payoffs. When lifespans are short ($M$ is small), agents experience greater uncertainty about which behavior is optimal given the fewer learning opportunities.
\end{enumerate}

@@

Model dynamics can be split into three parts: initialization,
\emph{intra}generational behavior and payoff accumulation, and
\emph{inter}generational transmission of the social learning trait and oblique
learning (Figure~\ref{fig:IntraInterGenerationalDynamics}). 
The model runs that we analyze here were initialized with $N=100$ agents
to randomly have the social learner trait or not, with no accumulated payoffs, and
blank ``ledgers''. At each time step, agents select one of $B$ possible behaviors
to perform, accumulating a payoff of 1 if the bandit pays off, and 0 if it does
not---a bandit pays off with probability $\pihigh$ if it is the unique optimal
behavior, or with probability $\pilow$ if it is one of $B-1$ non-optimal
behaviors. Agents accumulate payoffs in this way for $M$ time steps in each
generation. At the end of each generation, $N=100$ agents are randomly selected with
replacement to reproduce (asexual, haploid reproduction), 
weighted by their net payoffs over the generation,
i.e., \emph{performance-biased} reproduction. Child agents then learn from 
one teacher, selected again via a form of performance bias: a child agent chooses
its teacher by first selecting a fully random subset of agents from the 
parent generation as potential teachers, then selecting the one with the 
greatest net payoffs among the subset. This process continues until $T=1000$
time steps have elapsed \mt{this should probably be changed depending on $M$ so that
    all analyses use the same number of generations. my informal checks have shown quick
convergence, but i haven't tested b=4 and b=10 as much.}.

\begin{figure}
  \caption{Model dynamics are composed of two major parts: first, agents perform
  behaviors for $M$ time steps in a single generation. Then agents reproduce
  randomly, with more successful agents being more likely to reproduce. Newly
  ``born'' social learner agents use success bias to select and learn from a chosen
  teacher. Then, all agents from the previous generation die off, all new agents
  begin a new generation, and the process continues until 1000 total time steps
  have elapsed. Agents are initialized as ``blank slates''.}
  \label{fig:IntraInterGenerationalDynamics}
  \centering
    \includegraphics[width=0.9\textwidth]{Figures/IntraInterGenerationalDynamics.pdf}
\end{figure}

@@

The environment is initialized to have $B$ behaviors with one of them selected
at random to be the optimal behavior, denoted $b^*$. Behavior $b^*$ is 
initialized to have a probability $\pihigh$ of paying off 1. All other behaviors
are initialized to have probability $\pilow$ of paying off 1. The modeler must
specify the environmental variability, $u$, and the number of time steps per 
generation, $M$.

\vspace{2em}
\begin{table}[h]
    \caption{Global environmental and cognitive initialization parameters.}
    \label{tab:modelParameters}
    \centering \hspace{-3em}
    \begin{tabular}{cp{4.0in}l} \toprule

@@

        $M$    & Number of time steps per generation & 1, $B/2$, $B$, $2B$, $4B$ \\

        $N_T$    & Number of teachers to pool, from which best selected 
                 & \textbf{5} \mt{TODO: Test more for sensitivity analysis?} \\

        \bottomrule
        \end{tabular} 
\end{table}
\vspace{2em}

@@

Each generation begins with agents initialized either by the $t=0$ initialization
outlined above, or initialized via inter-generational reproduction and learning,
described in detail below. Within generations, there is no social learning.
At each time step within a generation, agents select a behavior to perform
using the softmax algorithm, then update their ledgers $\bar\pi_{ib}$ and behavior
counts $c_{ib}$. If 
the chosen bandit pays off for an agent, its net payoff is incremented by one,
$\pi'_i \leftarrow \pi_i + 1$. This process continues for $M$ time steps, 
when reproduction, learning, and die-off occur, re-initializing the next 
generation for performing $M$ behaviors selected via this same process.

At each time step, 
agent $i$ chooses behavior $b$ at random, with each behavior
weighted by the softmax function applied to that behavior's observed mean payoff
relative to all mean payoffs in the ledger,
\begin{equation}
  \Pr(i \text{ chooses behavior } b) = \frac{\exp(\bar\pi_{ib}
  / \tau) }{ \sum_{b=1}^B \exp(\bar\pi_{ib} / \tau)}.
\end{equation}  
Softmax behavior selection is a
biologically plausible model of behavior~\cite{Schulz2019} 
that enables agents to explore
alternative behaviors sometimes and exploit the best observed behavior other times,
in accordance with Luce's choice axiom~\cite{Luce1959},.
The parameter $\tau$ specifies how frequently alternative behaviors are
explored versus how frequently the best observed behaviors are 
exploited. Greater $\tau$ means more frequent exploration, lesser $\tau$ means
more frequent exploitation. We set $\tau = 0.1$ for
all computational analyses presented in the main paper. We (WILL HAVE) performed
sensitivity analyses and found a weak dependence on $\tau$ that does not affect our
main conclusions.

@@

Every $M$ time steps, agents from the current generation reproduce, teach,
and die, while the next generation to perform the next $M$ time steps learns
from one teacher from the previous generation if they inherited the social 
learner trait. Inter-generational dynamics thus depend on 
two payoff-biased selection mechanisms: reproducer selection
and teacher selection.

@@

\subsection{Computational analyses}

We manipulated environmental uncertainty parameters described above, $u$,
$\pisub{low}$, $B$, and $M$, to examine their effects on our
main outcome measure, the prevalence of social learning $s$. For each parameter setting
in our analysis we calculated the average value of $s$ at the final time step
of $T=1000$ across 100 runs for each combination of uncertainty parameter values.

To analyze the effect of various forms of uncertainty on social learning 
prevalence in our Analysis, we initialized several model runs with
different uncertainty parameters $u$, $\pisub{\mathrm{low}}$, $B$,
and $M$. We varied $u \in \{0.0, 0.1, \ldots, 1.0\}$ for each combination of
the following parameters:
$\pisub{low} \in \{0.1, 0.5, 0.8\}$, $B \in \{2, 4, 10\}$, $M \in \{B/2, B, 2B\}$.

\subsubsection{Outcome variables}

Our primary outcome variable is the mean prevalence of the social learning
trait over $N=100$ agents over 1000 trials at the final time step of each trial,
denoted $\meansl$.
Typically at the final time step all agents either have the social learning
trait or do not (see Supplement for a presentation of time series collections
and histograms of outcomes for
a representative sampling of variables \mt{TODO}).

\begin{table}[h]
    \caption{Global environmental and cognitive initialization parameters.}
    \label{tab:modelParameters}
    \centering \hspace{-3em}
    \begin{tabular}{cp{4.0in}l} \toprule

@@

        $p_{\text{opt}}$ & Fraction of population performing optimal behavior
                         & $\in [0.0, 1.0]$ \\

        \texttt{count}$(b_i)$ & Number of agents performing behavior $b$ 
                              & Behavior-count key-value pairs\\
        \bottomrule
    \end{tabular}
\end{table}

@@

We analyzed the outcomes of model runs across systematically varied uncertainty
parameters over 1000 trials for each parameter setting. 
We observed complex outcomes where the effects of uncertainty parameters
were highly dependent on one another as measured by the mean prevalence of
social learning, $\meansl$, across trials for each parameter setting. 
The prevalence of social learning
generally decreased with environmental variability, $u$, as expected from standard
models of social learning evolution. However, the shape of the decrease was not
uniform (Figure~\ref{fig:evolutionOfSL}). 
We observed that social learning did in fact enable populations to 
do better than individual learners, but only when (multi-dimensional) uncertainty 
were sufficiently limited (Figure~\ref{fig:meanNetPayoffs}).

\subsection{Evolution of social learning under uncertainty}
The shape of the decrease from all agents
always becoming social learners ($\meansl = 1.0$) to 
all agents always becoming individual learners ($\meansl = 0.0$)
is heterogeneous. We identified patterns in this heterogeneity that inform our
understanding of when and how social learning is advantageous, and when 
and why individual learning is advantageous. 
Changes in the uncertainty parameters ($\pilow$, $B$, $L$) 
result in the following changes in the shape of the transition from $\meansl = 1.0$ 
to $\meansl = 0.0$: (1) the transition
occurs at different critical values of $u$; and (2) the transition occurs
with greater or lesser rapidity over $u$. 
For instance, when $B=2$, $\pilow = 0.1$, and $L=8$, $\meansl$ goes from 1 to 
0 starting at $u=0.1$ and reaches 0 at $u=0.3$ (Figure~\ref{fig:evolutionOfSL},
top left). When we increased $B=4$ (Figure~\ref{fig:evolutionOfSL}, top center), the
transition from $\meansl = 1$ to 0 begins at $u=0.3$. When $B=4$, $\pilow=0.1$,
and $L=1$ the transition from $\meansl=1$ to 0 begins at $u=0.3$ but does not
reach 0 until $u \approx 0.9$ (again Figure~\ref{fig:evolutionOfSL}, top center).

We can draw some conclusions about the evolution of social learning by analyzing the
location and shape of the suppression of social learning as $u$ was increased.
First, let us observe some general trends. Increased $\pilow$ generally
flattened the suppression of social learning as $u$ increased.  
Increased $B$ tended to increase the critical value of $u$ at which the 
system transitions from favoring social learning to favor individual learning.  
Increased $L$ generally sharpened the suppression of social learning.
Within each of these broad trends over the uncertainty parameters are several
details that depend on context as defined by the values of the other uncertainty
parameters.

When $\pilow$ was increased, the suppression of social learning was, in
general, less rapid as $u$ increased. This effect is particularly acute when
$\pilow = 0.8$, which is very nearly equal to the expected value of the optimal
payoff, $\pihigh = 0.9$. When $\pilow = 0.8$ $\meansl$ never actually reaches 
1 or 0 for any parameter settings. The suppression of social learning in this
case is approximately linear, if there is selection at all. When $\pilow = 0.8$
and $L = 1$ there is little to no selection across $B$, 
so populations evolve randomly towards $\meansl = 1$ or $\meansl = 0$.

Increased $B$ generally led to increased critical $u$ at which social learning
starts to be suppresed. When payoffs are not ambiguous ($\pilow = 0.1$), the the
suppression of social learning starts at $u \approx 0.1$ for $B=2$; at $u \approx
0.3$ when $B=4$; and at $u \approx 0.5$ when $B=10$.  When payoffs are more
ambiguous ($\pilow = 0.45,0.8$), increased $B$ also led to a flattening of the
suppression of social learning when $L=1$.

The general effect of $L$ is to increase the rapidity of the suppression of social
learning via the strengthening of selection pressure since longer-lived agents
build up stronger priors over their lifetimes. If it is advantageous to
learn socially, this will be reliably detected. Inversely, decreased $L$
can weaken selection pressure since agents cannot amass useful information when
$L$ is small. This is most extreme when $\pilow = 0.8$ and there is little to
no selection for or against social learning when $L=1$. However, decreased $L$ can
also delay the suppression of social learning as $u$ increases, since social
learning provdes agents with the possibility of learning what the optimal behavior
might be.  Even when $u$ was large, agents with a lifespan of 1 more often evolved
to be social learners because social learning was to be preferred over random
chance.

\subsection{Social learning capacity increases payoffs}

For social learning to evolve, it must improve upon individual learning. 
We found that different parameter settings change how advantageous social learning
is over and above individual learning. To analyze the effect of uncertainty on
payoffs, it is useful to identify some constant reference values which we will
use to evaluate the benefit of social learning. First, the optimal expected payoff
is the optimal behavior payoff times the lifespan, which we can denote
$\langle \pi^*(L) \rangle = L\cdot\pihigh$. The expected base payoff by chance is 
$\langle \pi_{\text{base}} \rangle = L \left[ (B - 1) \pilow + \pihigh \right]$, which
an agent would receive on average from performing purely random behaviors. 
Agents learn adaptively over their lifespan (via the softmax algorithm in 
our model), whether or not they are social learners. We calculated this
value for reference for each triplet of $\pilow$, $B$, and $L$ 
\mt{calculate this and add to plot as a guide}. 
Social learning evolves when the expected payoff of
social learners exceeds the expected payoff of individual learners.

\mt{Use above framework to analyze results after plot guides added as
promised}

\subsection{Evolutionary outcomes vary in their certainty}

Finally, we found that the \emph{evolutionary uncertainty} of the system increases
around the values of $u$ where $\meansl$ goes from 1 to 0. 
In one sense
To confirm and expand our analysis of this uncertainty, we 
also measured environmental uncertainty in terms of the
expected number of time steps to convergence, $\langle T \rangle$. Evolutionary
uncertainty is an outcome measure---an emergent form of uncertainty due to
weak selection pressures. When 
$\meansl \approx 0.5$ this means that approximately equal numbers of trials
ended with all agents as social learners as they ended with all agents 
individual learners. We found that increased $\pilow$
tended to increase the overall uncertainty in terms of $\langle T \rangle$.
Increasing $B$ also increased the $u$ at which $\langle T \rangle$ peaked.
This indicates the evolutionary system is significantly less certain of 
which behavior is optimal around critical uncertainty parameter ensembles.

\mt{interpret this further}

\include{mainResultsPlots.tex}

\include{meanNetPayoffs.tex}

\include{stepResultsPlots.tex}

>>>>

\newcommand{\meanvar}[1]{\langle #1 \rangle}
\newcommand{\meansl}{\meanvar{s}}
\newcommand{\meanpi}{\meanvar{\pi}}
\newcommand{\meansoc}{\meanvar{\pi_\mathrm{S}}}
\newcommand{\meanasoc}{\meanvar{\pi_\mathrm{A}}}
\newcommand{\meanT}{\meanvar{T}}

@@

We developed an agent-based model of a society of $N$ individuals who each must decide
which of $B$ behaviors to perform at each time step. Each behavior is a ``bandit'',
a common modelling and experimental approach for representing behaviors with
probabilistic payoffs~\cite{SuttonBartoBook, McElreath2005,Rendell2010,Schulz2019}. %Bandits are like  slot machines in a casino~\ref{fig:behaviorSelection}.
Each behavior $b$ yields Bernoulli payoffs: payoff of 1 with probability $\pi_b$ and zero payoff
with probability $1 - \pi_b$; 
$\pi_b$ is therefore also the expected payoff of behavior $b$. Agents must
decide which behavior to perform at each time step. To do this, agents 
use an explore-exploit strategy to sometimes try the most profitable behavior
they know about, and other times try alternatives that may pay off more reliably.

We operationalized uncertainty in four different ways: 
(1) payoff ambiguity, $A_\pi$, which measures the difference
between the optimal expected payoff behavior $\pisub{high}$ and the expected payoff
of the other behaviors, $\pisub{low}$; 
(2) environmental variability, $u$, the probability the optimal behavior changes from one generation to another; 
(3) the number of possible behaviors, $B$; and 
(4) the lifespan, or time steps per generation, $L$.

We ran the simulation until agents evolved to be all social or all asocial 
learners, or until 20k time steps was reached. Only 44 out of 396k trials
did not reach fixation (all social or all asocial). At
the end of each generation agents reproduce and then die off. 
Those selected to reproduce
pass on their boolean social learning trait, $s$, without mutation,
which specifes whether agents learn socially. We developed a series of
computational analyses where we systematically vary the uncertainty
variables and observe the average prevalence of social learning trait $s$
across agents and trials, which is our primary outcome measure, 
denoted $\meansl$.

@@

In each time step, $N$ agents---autonomous problem solvers---select 
which behavior to perform based on their running tally of mean payoffs for 
each available behavior. %information (Figure~\ref{fig:behaviorSelection}). 
Agent $i$ tracks the mean payoff of each behavior $b$, denoted $\bar\pi_{ib}$, and
a \emph{count} of how many times it has performed each behavior, denoted $c_{ib}$.
$\bar\pi_{ib}$
is initialized to 0 for all $b$ at model initialization for
all agents, and after each generation for individual learner agents. Social
learning agents' $\bar\pi_{ib}$ is initialized as that of its teacher selected
through performance-biased oblique learning. Agent $i$'s accumulated payoffs
from performing several behaviors over their lifetime of $L$ time steps is
denoted $\pi_{i}$.

\begin{table}[h]
    \caption{Dynamic agent-level variables. Each has an implicit time dependence.}
    \label{tab:modelParameters}
    \centering %\hspace{-3em}
    \begin{tabular}{cp{2.5in}p{1.0in}} \toprule

@@

        $\pi_i$ & Net payoffs accumulated by $i$ within generation
                                & 0.0 \\
        \bottomrule
    \end{tabular}
\end{table}
\subsection{Modeling uncertainty}

In our model uncertainty is a tunable consequence of four environmental features:
\begin{enumerate}
    \item 
      We vary the frequency at which the optimal behavior changes, which we call
      \emph{environmental variability}, $u$. At the start of each generation, with
      probability $u$, a new behavior is assigned a payoff of $\pisub{high}$ while all
      other behaviors are assigned a payoff of $\pisub{low}$. Otherwise, the same
      behavior remains optimal across generations. 
    \item We vary the latent expected payoffs yielded by the bandits. For simplicity,
      we assume that in any given environmental state, there is one optimal behavior that
      yields an expected payoff of $\pisub{high}$, while all other behaviors yield a 
      payoff of $\pisub{low} < \pisub{high}$. As $\pilow$ increases towards 
      $\pihigh$, uncertainty in the form of \emph{payoff ambiguity} also increases,
      making it more difficult, and less rewarding, for agents to determine which
      behavior is the optimal one.
    \item We vary the total \emph{number of available behaviors}, $B$, which is a
      source of uncertainty since agents are less likely to know which behavior yields high payoffs.  
    \item Finally, we vary number of behavioral events per generation, $L$. This can be viewed as the \emph{effective lifespan} of an agent.  Decreasing this lifespan effectively increases the importance of each event for acquiring payoffs. When lifespans are short ($L$ is small), agents experience greater uncertainty about which behavior is optimal given the fewer learning opportunities.
\end{enumerate}

@@

Model dynamics can be split into three parts: initialization,
\emph{intra}generational behavior and payoff accumulation, and
\emph{inter}generational transmission of the social learning trait and oblique
learning (Figure~\ref{fig:IntraInterGenerationalDynamics}). 
The model runs that we analyze here were initialized with $N=100$ agents
to randomly have the social learner trait or not, with no accumulated payoffs, and
blank ``ledgers''. At each time step, agents select one of $B$ possible behaviors
to perform, accumulating a payoff of 1 if the bandit pays off, and 0 if it does
not---a bandit pays off with probability $\pihigh$ if it is the unique optimal
behavior, or with probability $\pilow$ if it is one of $B-1$ non-optimal
behaviors. Agents accumulate payoffs in this way for $L$ time steps in each
generation. At the end of each generation, $N=100$ agents are randomly selected with
replacement to reproduce (asexual, haploid reproduction), 
weighted by their net payoffs over the generation,
i.e., \emph{performance-biased} reproduction. Child agents then learn from 
one teacher, selected again via a form of performance bias: a child agent chooses
its teacher by first selecting a fully random subset of agents from the 
parent generation as potential teachers, then selecting the one with the 
greatest net payoffs among the subset. This process continues until agents
have become all social or all asocial learners, or when $20k$ time steps have
elapsed.

\begin{figure}
  \caption{Model dynamics are composed of two major parts: first, agents perform
  behaviors for $L$ time steps in a single generation. Then agents reproduce
  randomly, with more successful agents being more likely to reproduce. Newly
  ``born'' social learner agents use success bias to select and learn from a chosen
  teacher. Then, all agents from the previous generation die off, all new agents
  begin a new generation, and the process continues until 1000 total time steps
  have elapsed. Agents are initialized as ``blank slates''. \mt{Can compress this
  into triptych. $M \rightarrow L$. Lighten grey background, remove outlines
around most or all components. Replace text in picture with A, B, C, and 
explain in caption.}}
  \label{fig:IntraInterGenerationalDynamics}
  \centering
    \includegraphics[width=0.9\textwidth]{Figures/IntraInterGenerationalDynamics.pdf}
\end{figure}

@@

The environment is initialized to have $B$ behaviors with one of them selected
at random to be the optimal behavior, denoted $b^*$. Behavior $b^*$ is 
initialized to have a probability $\pihigh$ of paying off 1. All other behaviors
are initialized to have probability $\pilow$ of paying off 1. The modeler must
specify the environmental variability, $u$, and the number of time steps per 
generation, $L$.

\vspace{2em}
\begin{table}[h]
    \caption{Global environmental and cognitive initialization parameters.
    \mt{TODO: Test more $N_\text{T}$ for sensitivity analysis}}
    \label{tab:modelParameters}
    \centering %\hspace{-3em}
    \begin{tabular}{cp{2.5in}p{1in}} \toprule

@@

        $L$    & Number of time steps per generation & 1, $B/2$, $B$, $2B$, $4B$ \\

        $N_T$    & Number of teachers to pool, from which best selected 
                 & \textbf{5}  \\

        \bottomrule
        \end{tabular} 
\end{table}

@@

Each generation begins with agents initialized either by the $t=0$ initialization
outlined above, or initialized via inter-generational reproduction and learning,
described in detail below. Within generations, there is no social learning.
At each time step within a generation, agents select a behavior to perform
using the softmax algorithm, then update their ledgers $\bar\pi_{ib}$ and behavior
counts $c_{ib}$. If 
the chosen bandit pays off for an agent, its net payoff is incremented by one,
$\pi'_i \leftarrow \pi_i + 1$. This process continues for $L$ time steps, 
at which point reproduction, learning, and die-off occur, which re-initializes 
the next generation for performing $L$ behaviors selected via this same process.

At each time step, 
agent $i$ chooses behavior $b$ at random, with each behavior
weighted by the softmax function applied to that behavior's observed mean payoff
relative to all mean payoffs in the ledger,
\begin{equation}
  \Pr(i \text{ chooses behavior } b) = \frac{\exp(\bar\pi_{ib}
  / \tau) }{ \sum_{b=1}^B \exp(\bar\pi_{ib} / \tau)}.
\end{equation}  
Softmax behavior selection is a
biologically plausible model of behavior~\cite{Schulz2019} 
that enables agents to explore
alternative behaviors sometimes and exploit the best observed behavior other times,
in accordance with Luce's choice axiom~\cite{Luce1959},.
The parameter $\tau$ specifies how frequently alternative behaviors are
explored versus how frequently the best observed behaviors are 
exploited. Greater $\tau$ means more frequent exploration, lesser $\tau$ means
more frequent exploitation. We set $\tau = 0.1$ for
all computational analyses presented in the main paper. We 
 performed
sensitivity analyses and found a weak dependence on $\tau$ that does not affect our
main conclusions \mt{have run $\tau$ sensitivity checks, need to analyze them}.

@@

Every $L$ time steps, agents from the current generation reproduce, teach,
and die, while the next generation to perform the next $L$ time steps learns
from one teacher from the previous generation if they inherited the social 
learner trait. Inter-generational dynamics thus depend on 
two payoff-biased selection mechanisms: reproducer selection
and teacher selection.

@@

\subsection{Computational analyses and outcome measures}

We manipulated environmental uncertainty parameters described above, $u$,
$\pisub{low}$, $B$, and $L$, to examine their effects on our
main outcome measure, the mean prevalence of social learning $\meansl$, 
and supplemental outcome measures $\meanpi / L$ and $\meanT / L$. $\meanpi / L$ is the
mean net payoffs accumulated by agents at the end of each generation for the
final generation of agents, normalized by lifespan $L$. $\meanT / L$ is the 
mean time steps to convergence normalized by the lifespan of agents, i.e., the
mean number of generations until convergence was reached. Mean values were
caluculated across 100 model runs for each combination of uncertainty parameter values.
To test the effect of the uncertainty parameters, 
we initialized several model runs with
systematically varied uncertainty parameters, $u$, $\pisub{\mathrm{low}}$, $B$,
and $L$. We varied $u \in \{0.0, 0.1, \ldots, 1.0\}$ for each combination of
the following parameters:
$\pisub{low} \in \{0.1, 0.5, 0.8\}$, $B \in \{2, 4, 10\}$, and
$L \in \{1, B, 2B, 4B\}$ for $B=2$ and $L \in \{1, B/2, B, 2B\}$ for $B=4,10$.

\begin{table}[h]
    \caption{Outcome variables}
    \label{tab:outcomeVariables}
    \centering %\hspace{-3em}
    \begin{tabular}{cp{2.5in}p{1in}} \toprule

@@

        $\meanpi / L$ & Mean payoffs accumulated in a generation normalized by
        lifespan & $\in [0.0, 1.0]$ \\

        $\meanT / L$ & Mean number of generations to convergence & Bounded above by 20k / 8\\
        \bottomrule
    \end{tabular}
\end{table}

\subsection{A note on relationship with standard RL formalism and models}

We presented our model as an agent-based model because that is
how we originally developed it. Howevever, we believe the model could be 
translated in a straightforward way where each action at time $t$, typically denoted
$a_t$, is to ``pull'' one of the $B$ bandits, and the policy $\pi$ would 
involve using the softmax function to weight random behavior
selection~\cite{SuttonBartoBook}. Because of the complication of two
relevant time intervals, intervals between actions and generational intervals, and
space limitations we do not try to put the entire model in standard RL terms.
\mt{Expand/improve.}

@@

To understand the effect of uncertainty parameters (1) environmental variability, $u$; (2) payoff
ambiguity that increases with $\pilow$; (3) number of behavioral options, $B$; and 
(4) generation length/agent lifespan, $L$, on the evolution of social learning, we analyze the results of
model runs across systematically varied parameter combinations.  We observed complex
outcomes where the effects of uncertainty parameters were highly dependent on one
another as measured by the mean prevalence of social learning, $\meansl$, across
trials for each parameter setting.  Increased environmental variability, $u$, 
tended to suppress social learning as expected from standard models of
social learning evolution. However, the shape of the decrease in $\meansl$ 
was not uniform (Figure~\ref{fig:evolutionOfSL}). 
Depending on the parameter settings, the critical value of $u$ at which social 
learning suppression began varied, as did the steepness of the decrease. 
We found that the location and width of the decrease in $\meansl$ is correlated
with the difference between the expected net payoffs of a population of all
social learners, $\meansoc$, and the expected net payoffs
of an all-asocial learner population, $\meanasoc$ (Figure~\ref{fig:meanNetPayoffs}). 
evolution does not always choose the theoretically highest payoff strategy,
and confirmed that evolution is more uncertain (i.e.\ selection is
weaker~\cite[p. 103]{BoydRicherson1985}) when $\meansoc - \meanasoc$ is small, 
where evolutionary uncertainty is measured by the 
average steps to model convergence, $\meanvar{T}$ (Figure~\ref{fig:steps}).

\subsection{Evolution of social learning under uncertainty}

We observed that different uncertainty contexts lead to different patterns in
the evolution and suppression of social learning. To understand the evolution
and suppression of social learning we inspected the social learning prevalence,
$\meansl$, which is the average of the frequency of social learning across
a model population over all trial model runs for a 4-tuple of uncertainty
values, $(u, \pilow, B, L)$.  We plot $\meansl$ on the y-axis 
over $u=0.0,0.1,\ldots,1.0$ on the x-axis in each plot, with nine plots total over
$\pilow=0.1,0.45,0.8$ and $B=2,4,10$, broken out by $L=1,B/2,B,2B$ within each plot
(Figure~\ref{fig:evolutionOfSL}). Across uncertainty scenarios, $\meansl$ monotonically
decreases $u$ increases, in accordance with classic cultural evolutionary
predictions~\cite{CavalliFeldman1981,BoydRicherson1985,Feldman1996}.  Therefore we compare
differences between uncertainty contexts in terms of the \emph{suppression of social learning}
over (increasing) $u$.

Differences emerged in the suppression of social learning on two dimensions.
First there were systematic differences in the value of $u$ when social
learning first begins to be suppressed, i.e., where $\meansl$ first decreases. We show
in the following subsection that this occurs when $\meansoc$ approaches $\meanasoc$.
The second dimension of analysis is the width of the transition from 
maximal to minimal $\meansl$. In some cases, this width was less than 1, but
in many other cases the width was equal to 1, meaning that $\meansl$ 
decreased over all $u$. This indicated weak selection pressure since 
less than 0.001\% of all trials ended with all agents either social or asocial
learners. This is confirmed below by examining the average number of time steps
it took to reach fixation (agents all social or all asocial).

For now we will examine differences in the location and width of social 
learning suppression. \mt{In outline form while I work out each point}
\begin{itemize}
  \item 
    First, note that as $\pilow$ increases towards $\pihigh = 0.9$, social learning
    suppression widens, becoming linearly decreasing over $u$ with no $\meansl$
    reaching 1 or 0. This is due to weakened selection as the difference between
    social and asocial learning payoffs goes to zero.
  \item
    Second, as $B$ increases, note that social learning suppression 
    begins at increasingly high $u$. This is because it is increasingly worth
    the risk that a socially-learned ledger biases agents towards non-optimal
    behaviors when the number of choices increases. Furthermore, when $B$ is
    larger, one large ledger entry is not as misleading as when $B$ is smaller
    since there is more probability mass in the softmax vector's non-maximal
    behavior entries when $B$ is larger, and so alternatives to the socially
    learned ``best'' behavior will be explored more frequently.
  \item
    Finally, note that decreased lifespan both increases the location of 
    social learning suppression and the width of social learning suppression,
    with the extreme case of $L=1$ becoming linear and eventually flat compared
    to other values of $L$ as $\pilow$ and $B$ increase.
\end{itemize}
To support these explanations we need to analyze the average payoffs 
obtained by agents in these various uncertainty contexts and compare them
with the expected payoffs when all agents are individual learners, and with the
expected payoffs when all agents are social learners.

\subsection{Relative benefit of social learning}

Social learning begins to be suppressed when the expected payoffs from 
a population of all social learners, $\meansoc$ (Figure~\ref{fig:meanNetPayoffs};
small dashed line with diamond markers),
decreased to the point where they are nearly the same as expected payoffs for
a population of all individual learners, $\meanasoc$
(Figure~\ref{fig:meanNetPayoffs}; horizontal long-dash lines). We see that social 
learning becomes completely suppressed when $\meansoc < \meanasoc$.  Mean 
payoffs realized by model agents approximately track expected payoffs of
whichever strategy is optimal, social or asocial learning, but with some important
variation.

Similarly, sometimes populations that start out mixed outperform all
social learner populations when $\meansoc > \meanasoc$, indicating that an initial
diversity of learner types helps with the long-term prosperity of the group, even
though the asocial learner type tends to die out. \mt{I suspect
this may be sensitive to $N_T$, so my next analysis will be the effect
of $N_T$.}

Second, notice that asocial learning may be the safe fallback option when 
evolution is unsure which strategy is optimal, 
but a population of all social learners would do better
than what evolution tended to select for (Figure~\ref{fig:meanNetPayoffs},
e.g., when $\pilow=0.45$, $B=2$, and $L=8$ in the middle left plot). \mt{I suspect
this will also be sensitive to $N_T$.}

\subsection{Selection strength and evolutionary uncertainty}

If selection is truly weaker when $\meansoc \approx \meanasoc$, then we should not
only see a more perfect bimodality of outcomes, with roughly half of outcomes going
to both $\meansl = 1$ and $\meansl = 0$, but we should also see the models taking
longer to converge in this situation.  Indeed, weak selection is also evidenced by
spiking or elevated average time to convergence, $\meanT$, around when theoretical
$\meansoc$ and observed $\meanpi$ approach $\meanasoc$. Recall that models are run
until they reach fixation or until they reach 20k time steps. Only 44 out of 396k
did not reach fixation, less than 0.001\% of all trials.

\input{mainResultsPlots.tex}

\input{meanNetPayoffs.tex}

\input{stepResultsPlots.tex}
